---
title: "HW3"
author: "Ryan Gallagher"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
data('iris')
library(gtsummary)
library(tidyverse)
```

This is a simple example of automatic report generation from R. Analyses were performed using R version 4.3.1 (2023-06-16 ucrt)

## Analysis Methods

Fisher's iris data set that is available as `iris` within R is analyzed. The 150 flower measurements were summarized using mean and standard deviation within each species, and compares between groups using ANOVA.

```{r echo=FALSE}
iris %>% select(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width, Species) %>%
  tbl_summary(by = Species, 
              statistic=all_continuous() ~ "{mean} ({sd})") %>%
  add_p(test = everything() ~ "aov") %>%
  add_overall()
```

## Relationship between petal length and width
```{r echo=FALSE}
ggplot(data=iris, aes(x=Petal.Length, y=Petal.Width)) + 
  geom_point() + 
  stat_smooth(method='lm',
              formula= y ~ x,
              geom="smooth") +
  labs(x="Petal length, mm", y="Petal width, mm")
```
\
Figure 1: Scatterplot of petal width vs petal length 
\
Simple linear regression was used to evaluate the relationship between petal length and petal width using the model:
$$\text{Petal Width} = \alpha + \beta(\text{Petal Length}) + \epsilon$$
where $\epsilon$ ~ $N(0, \sigma^2)$.
\

<center> 
#### Table 1. \
Simple Linear Regression estimates predicting petal width based on petal length across all species 
</center>

```{r echo=FALSE}
reg = glm(Petal.Width ~ Petal.Length, data=iris) 
reg %>% tbl_regression(intercept=1)
```
Based on the regression model, we estimate that petal length explaines 92.7% of variability in petal width.