---
title: "Chapter8_HW"
author: "Ryan Gallagher"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 3. Consider the Gini index, classification error, and entropy in a single classification setting with two classes. Create a single plot that displays each of these quantiles as a function of $\hat{p}_{m1}$. The x-axis should display $\hat{p}_{m1}$, ranging from 0 to 1, and the y-axis should display the value of the Gini index, classification error, and entropy.

```{r}
library(ggplot2)
#sim data
p = seq(0, 1, length.out = 500)

#pm1 = 1-pm2

# calculate values
gini = 1 - (p^2) - ((1 - p)^2)
classification_error = 1 - pmax(p, 1 - p)
entropy = - (p * ifelse(p == 0, 0, log2(p)) + (1 - p) * ifelse((1 - p) == 0, 0, log2(1 - p)))
data = data.frame(p, gini, classification_error, entropy)

#plot it
ggplot(data, aes(x = p)) +
  geom_line(aes(y = gini, color = "Gini Index")) +
  geom_line(aes(y = classification_error, color = "Classification Error")) +
  geom_line(aes(y = entropy, color = "Entropy")) +
  labs(title = "Gini Index, Classification Error, and Entropy vs Probability",
       x = expression(hat(p)[m1]),
       y = "Measure Value") +
  theme_minimal() +
  scale_color_manual(values = c("Gini Index" = "red", "Classification Error" = "blue", "Entropy" = "green"))

```

## 7. In the lab, we applied random forests to the *Boston* data using $mtry=6$ and using $ntree=25$ and $ntree=500$. Create a plot displaying the test error resulting from random forests on this data set for a more comprehensive range of values for $mtry$ and $ntree$. You can model your plot after figure 8.10. Describe the results obtained.

8.10 has the y-axis as the Test Classification Error and the x-axis as the number of trees (ntree). There are lines colored by the value of m (mtry).

```{r}
library(ISLR2)
library(randomForest)

set.seed(1)
train = sample(1:nrow(Boston), nrow(Boston) / 2)
boston.test = Boston$medv[-train]

ntree_seq = seq(25,500, by=25)
mtry_seq = 1:12

results = data.frame(ntree=integer(), mtry=integer(), test_error=numeric())

for(ntree in ntree_seq){
  for (mtry in mtry_seq){
    set.seed(1)
    rf.boston = randomForest(medv ~ ., data = Boston, 
                              mtry = mtry,
                              ntree = ntree, 
                              subset = train)
    
    yhat.bag = predict(rf.boston, 
                        newdata = Boston[-train, ])
    
    test_error = mean((yhat.bag - boston.test)^2)
    
    # store results
    results = rbind(results, data.frame(ntree, mtry, test_error))
  }
}

ggplot(results, aes(x = ntree, y = test_error, color = as.factor(mtry))) +
  geom_line() +
  labs(title = "Test Error vs. Number of Trees",
       x = "Number of Trees (ntree)",
       y = "Test Error",
       color = "mtry") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5))
```

## 9. This problem involves the $OJ$ data set which is a part of the $ISLR2$ package.

### (a) Create a training set containing a random sample of 800 observations, and a test set containing the remaining observations.

```{r}
library(ISLR2)
library(tree)
train_idx = sample(1:nrow(OJ), 800)
train = OJ[train_idx, ]
test = OJ[-train_idx, ]
```

### (b) Fit a tree to the training data, with *Purchase* as the response and the other variables as predictors. Use the *summary()* function to produce summary statistics about the tree, and describe the results obtained. What is the training error rate? How many terminal nodes does the tree have?

```{r}
tree.train = tree(Purchase~., data=train)
summary(tree.train)
```

This tree has a training error rate of 0.1762 with 9 terminal nodes.

### (c) Type in the name of the tree object in order to get a detailed text output. Pick one of the terminal nodes and interpret the information displayed.

```{r}
tree.train
```

Well the initial node splits the purchase decision based on how loyal the customer is to Citrus Hill. Looking at all the outcomes for those with a loyalty value less than 0.48285, they always choose Minute Maid. For more loyal Citrus Hill drinkers, there is a decision made based on the price difference between CH and MM, as well as any discount considerations. What this tells me is that a customer would have to be quite loyal to CH to pick it over MM.

### (d) Create a plot of the tree, and interpret the results

```{r}
plot(tree.train)
text(tree.train, pretty=0)
```

### (e) Predict the response on the test data, and produce a confusion matrix comparing the test labels and the predicted test labels. What is the test error rate?

```{r}
tree.test = predict(tree.train, test,
                    type = 'class')
table(tree.test, test$Purchase)

(140+92) / 270
```

The test error rate is $0.85925$. Pretty good!

### (f) Apply the *cv.tree* function to the training set in order to determine the optimal tree size.

```{r}
set.seed(2)
cv.OJ = cv.tree(tree.train, FUN = prune.misclass)
names(cv.OJ)
cv.OJ
```

We find that 9,6, and 4 all have very close deviation, but 9 has the smallest.

### (g) Produce a plot with tree size on the x-axis and cross-validated classification error rate on the y-axis

```{r}
par(mfrow = c(1, 2))
plot(cv.OJ$size, cv.OJ$dev, type='b')
plot(cv.OJ$k, cv.OJ$dev, type='b')
```

### (h) Which size tree corresponds to the lowest cross-validated classification error rate?

A tree size of 9 produces the lowest cross-validated classification error rate.

### (i) Produce a pruned tree corresponding to the optimal tree size obtained using cross-validation. If cross-validation does not lead to selection of a pruned tree, then create a pruned tree with 5 terminal nodes.

I already plotted the 9 node tree. So this doesn't prune anything. I'll try 4 even though 9 has a smaller error rate.

```{r}
prune.OJ = prune.misclass(tree.train, best=4)
plot(prune.OJ)
text(prune.OJ, pretty=0)
```

### (j) Compare the training error rates between the pruned and unpruned trees. Which is higher?

```{r}
summary(prune.OJ)
```

The unpruned tree has a smaller error rate of

```         
0.1762 = 141 / 800
```

Where our pruned tree has an error rate slightly larger at 0.18 = 144/800.

### (k) Compare the test error rates between the pruned and unpruned trees. Which is higher?

```{r}
prune.test = predict(prune.OJ, test,
                    type = 'class')
table(prune.test, test$Purchase)


```

The nonpruned tree had a test error rate of:

```         
0.8592593
```

This pruned tree has a test error rate of

```{r}
(144 + 88) / (144+88+16+22)
```

Which is exactly the same as before.

## 10. We now use boosting to predict *Salary* in the *Hitters* data set.

### (a) Remove the observations for whom the salary information is unknown, and then log-transform the salaries

```{r}
library(tidyverse)
Hitters = Hitters %>% filter(is.na(Salary) == FALSE) %>% 
  mutate(Salary = log(Salary))
```

### (b) Create a training set of consisting of the first 200 observations, and a test set consisting of the remaining observations

```{r}
train_idx = sample(1:nrow(Hitters), 200)
train = Hitters[train_idx, ]
test = Hitters[-train_idx, ]
```

### (c) Perform boosting on the training set with 1000 trees for a range of values of the shrinkage parameter $\lambda$. Produce a plot with different shrinkage values on the x-axis and the corresponding training set MSE on the y-axis.

```{r}
library(gbm)
library(ggplot2)

set.seed(1)

shrinkage_values = seq(0.01, 0.5, by = 0.05)
results = data.frame(shrinkage = numeric(), MSE = numeric())

for(lambda in shrinkage_values) {
  boost.hitters = gbm(Salary~., data = train,
                       distribution = 'gaussian', n.trees = 1000,
                       shrinkage = lambda)


  pred = predict(boost.hitters, n.trees = 1000)
  mse = mean((pred - train$Salary)^2)
  results = rbind(results, data.frame(shrinkage = lambda, MSE = mse))
}
ggplot(results, aes(x = shrinkage, y = MSE)) +
  geom_line() +
  labs(title = "Training MSE vs. Shrinkage",
       x = "Shrinkage",
       y = "Training MSE") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust=0.5))

```

### (d) Produce a plot with different shrinkage values on the x-axis and corresponding test set MSE on the y-axis

```{r}
set.seed(1)

shrinkage_values = seq(0.01, 0.5, by = 0.05)
results = data.frame(shrinkage = numeric(), test_MSE = numeric())

for(lambda in shrinkage_values) {
  boost.hitters = gbm(Salary~., data = train,
                       distribution = 'gaussian', n.trees = 1000,
                       shrinkage = lambda)


  pred = predict(boost.hitters, newdata = test, n.trees = 1000)
  test_mse = mean((pred - test$Salary)^2)
  results = rbind(results, data.frame(shrinkage = lambda, test_MSE = test_mse))
}

ggplot(results, aes(x = shrinkage, y = test_MSE)) +
  geom_line() +
  labs(title = "Test MSE vs. Shrinkage",
       x = "Shrinkage",
       y = "Test MSE") +
  theme_minimal() + 
  theme(plot.title=element_text(hjust=0.5))

```

### (e) Compare the test MSE of boosting to the test MSE of that results from applying two of regression approaches seen in Chapters 3 and 6.

Linear Regression

```{r}
lm.model = lm(Salary~., train)
lm.preds = predict(lm.model, test)
lm.test.err = mean((lm.preds - test$Salary)^2)
lm.test.err
```

Lasso Regression

```{r}
library(glmnet)
x.model.mat = model.matrix(Salary~., train)
y.model.mat = model.matrix(Salary~., test)

y = train$Salary

lasso.model = glmnet(x.model.mat, y, alpha = 1)
lasso.preds = predict(lasso.model, s = 0.01, y.model.mat)
lasso.test.err = mean((lasso.preds - test$Salary)^2)

lasso.test.err
```

The test MSE of the boosting is:

```{r}
min(results$test_MSE)
```

So we find that the boosting test MSE is **much** smaller than the other methods.

### (f) Which variables appear to be the most important predictors in the boosting model?

```{r}
boosted.model = gbm(Salary~., data = train, 
                              distribution = "gaussian", 
                              n.trees = 1000, 
                              shrinkage = 0.01)

summary(boosted.model)
```

The most important variables seem to be CRuns, CAtBat, and Chits.

### (g) Now apply bagging to the training set. What is the test MSE for this approach?

```{r}
bag.model = randomForest(Salary~., data = train, 
                              distribution = "gaussian", 
                              n.trees = 1000, 
                              shrinkage = 0.01,
                              mtry = 19,
                              importance = TRUE)



bag.preds = predict(bag.model, test)

bag.test.err = mean((bag.preds - test$Salary)^2)

bag.test.err
```

We get a test MSE of 0.2256. Better than the boosting method above by using this method which takes all predictors into account .
